A scoping review of rodent trapping studies to understand the prevalence of zoonotic disease in West Africa.

### 2021-10-04 reports/manuscript_v2.Rmd
clean merge

---
title: " A scoping review of rodent trapping studies to understand the prevalence of zoonotic disease in West Africa."
description: |
date: "`r Sys.Date()`"
author: 
  - first_name: "David"
    last_name: "Simons"
    url: https://www.dsimons.org
    orcid_id: 0000-0001-9655-1656
output: 
  distill::distill_article:
    toc: true
    toc_depth: 3
    toc_float: true
  # word_document:
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_dir = here::here())
  })
always_allow_html: yes
bibliography: "`r here::here('citations', 'manuscript_citations.bib')`"
---
```{r, echo=FALSE}
source(here::here("scripts", "manuscript_library.R"))
```

@@

Rodents are important reservoirs of zoonotic infectious diseases; 255 species are known hosts of 85 zoonotic pathogens. These reservoir species are globally distributed with West Africa containing multiple known and several further predicted reservoir species. Typically International Union for Conservation of Nature (IUCN) species distribution maps and Global Biodiversity Information Facility (GBIF) presence maps are used to determine regions at risk for zoonotic spillover events, however, these datasets are biased by incomplete sampling. To investigate the impact of rodent species sampling heterogeneity on zoonotic spillover risk we have systematically reviewed rodent trapping studies to produce a contextually rich dataset that can be used to explore this risk at finer spatial resolution. Here we show that sampling of rodents and their pathogens in the West African region are spatially biased by country and trapping habitat. We found that rodent trapping effort was associated with regional population density and was increased in habitats modified by human activity. We produce updated rodent species ranges compared to available IUCN maps and enrich GBIF data by including locations of rodent species absence and trapping effort. Furthermore, we report the spatial bias in the investigation of four important rodent zoonoses, Arenaviridae, *Borellia sp.*, *Bartonella sp.* and *Toxoplasma sp.*. Our results highlight that incorporating sampling bias is important when assessing the risk for zoonotic spillover events from rodents. The synthesis of contextually rich rodent trapping data contributes important information that is lacking in IUCN distribution maps and GBIF species presence data. We anticipate this dataset can support the production of more complete spatial risk assessments of zoonotic spillover events. For example, the inclusion of absence data and trapping effort can help identify regions where data sparsity may produce inappropriately predicted risk. Furthermore, this data specifies regions in need of comprehensive study for four targeted rodent zoonoses, Arenaviridae, *Borellia sp.*, *Bartonella sp.* and *Toxoplasma sp.*.

@@

The potential effect of zoonotic infectious diseases outbreaks on human health and societies has been dramatically highlighted through the ongoing SARS-CoV-2 pandemic and recent Ebola virus outbreaks. The number of zoonotic spillover events are projected to increase under intensifying anthropogenic pressure such as, increased human populations [@Allen], increasing urbanisation [@Hassell], and global climate change [@morse_prediction_2012]. In addition, two taxa are proposed to contribute to the highest burden of potential zoonoses globally, rodents (Rodentia) and bats (Chiroptera) [@han]. Of 2,220 extant rodent species, 244 (10.7%) are described as being reservoirs of 85 zoonotic pathogens [@han].  Specifically, West Africa has been identified as region at increased hazard for rodent-borne zoonotic spillover events under different projected scenarios [@garcia-pena_land-use_2021]. Rodents are implicated in the transmission pathways of several diseases in this region, including, Lassa fever, Schistosomiasis and Leptospirosis.

Rodent species form diverse assemblages providing important ecosystem services such as pest regulation and seed dispersal [@fischer_ecosystem_2018]. However, the role a minority of species play in zoonotic infectious disease spillover can be considered an ecosystem disservice. Rodents typically demonstrate “fast” life histories [@dobson_fast_2007], these traits are also associated with a species being a reservoir of zoonotic pathogens [@han_rodent_2015; @albery_fast-lived_2021]. Further, these traits are prevalent in species that thrive in human dominated landscapes where they displace species that are less likely to be reservoirs of zoonotic disease [@gibb_zoonotic_2020]. The widespread occurrence of reservoir species and their proximity to human activity make the description of rodent species assemblages and host-pathogen interactions vitally important.

Rodent trapping studies provide a useful method to describe rodent population structures and survey for potential zoonoses. Studies have been conducted in West Africa to both identify novel potential zoonotic pathogens within rodents [@predicts] and to investigate the prevalence and burden of known pathogens within known rodent hosts (e.g. for Lassa fever [@fichet-calvet_diversity_2009] and Schistosomiasis [@catalano_multihost_2020]). These studies provide contextually rich information on when, where and under what conditions rodents were trapped that are typically missing from the global datasets [@bovendorp_optimising_2017]. Despite this, these global datasets (e.g. IUCN, GBIF and GIDEON) are typically used to inform species distribution maps and zoonotic disease spillover predictions [@han_global_2016; @smith_global_2014; @pigott_mapping_2014].
.
These studies have been used to identify potential geographic hotspots where virus and host species diversity may be expected to be at its greatest to predict regional zoonotic disease spillover risks. However, there remains the potential for important confounding in these spatial distributions through bias generated by study design and selection of sampling sites. For example, systematically increased sampling (e.g. more intensive studies over longer time periods) or over-representation of certain habitats (e.g. periurban landscapes) could lead to an apparent association between locations and risk that is driven by these factors rather than an underlying host and virus association [@wille_how_2021; @gibb_mammal_2021]. Conversely some regions may not be sampled adequately and therefore under-represented in these datasets due to sparse human populations or inaccessible habitats which may lead to proposing that these regions are at low risk of zoonotic disease spillover events.

@@

For the subset of studies investigating rodent zoonoses we compared the location of trapping sites with SEDAC Global Population Density estimates for 2005, the median year studies were commenced. We used a Generalised Additive Model (GAM) incorporating a spatial interaction term to investigate the association of number of trap nights and human population density. The model structure was specified as:

@@

4,282 relevant citations were identified, with 126 rodent trapping studies included in narrative synthesis. The earliest included studies were published in 1974 with increasing numbers of studies being performed annually since 2005. The median length of rodent trapping activity was one year (IQR 0-2 years). The median time from completion of rodent trapping to publication is 3 years (IQR 2-5 years) (Figure 1.).

@@

knitr::include_graphics(here("figures", "Fig_1_Panel_A.png"))

@@

Rodent trapping took place at 1,331 trap sites with at least one trap site recorded from 14 West African countries. No rodent trapping studies were identified from Gambia or Togo. Thirty-one (25%) studies reported trapping at a single study site, 46 (37%) studies trapped at between two and five study sites, the remaining 49 studies trapped at between six and 93 study sites. Trap sites were situated in 273 (19.3%) of 1414 level 2 administrative regions from the 14 West African nations. The areas with highest trapping activity included the capital cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako) and Ghana (Accra) and the largest cities of Côte d'Ivoire (Abidjan) and Benin (Cotonou). Outside of these cities, Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone were the most intensively studied regions (Figure 2. A).

Greater numbers of trap nights were conducted in areas with higher population densities (*R~2~* = 0.167). In particular Mauritania, Northern Senegal, Eastern Sierra Leone, Eastern Guinea and South West Nigeria were trapped at increased rates than would be expected based on their population densities. South West Nigeria, Northern Nigeria, Liberia, Côte d'Ivoire, Ghana, Niger and Burkina Faso were trapped at lower rates (Figure 2. B) .

@@

```{r, figure_2, layout = "l-screen-inset", echo=FALSE}
knitr::include_graphics(here("figures", "Figure_2.png"))
```

@@

73,164 small mammals were trapped (592 were trapped outside of West African countries), 2,830 (4%) trapped individuals were identified to order level (Rodentia), 7,760 (11%) were identified to genus level, the remaining 62,574 (85%) were identified to species level. In studies reporting the number of trap nights the median trap success rate of traps placed in or around buildings was 13% (IQR 6-24%), this compares to a median trap success rate of 3% (IQR 1-9%) in other habitats. Of the 147 distinct species trapped (see Supplementary Material 3) the majority were from the order Rodentia (112). Muridae (82) represented the largest family of rodents, followed by, Sciuridae (10), Gliridae and Nesomyidae (both 6), Ctenodactylidae and Anomaluridae (2) and Dipodidae, Hystricidae and Thryonomidae (all 1). The remaining 34 species came from the orders of the Soricomorpha (30), Erinaceomorpha (2) and Afrosoricida (1).

The most commonly trapped genera of rodents were *Mastomys sp.* (27,079, 37.7%), *Rattus sp.* (11,472, 16.1%), *Mus sp.* (8,624, 12.0%), *Arvicanthis sp.* (5,821, 8.1%) and *Praomys sp.* (5,409, 7.5%). At species level *Mastomys natalensis* (11,116, 17.4%), *Rattus rattus* (9,959, 15.6%), *Mastomys erythroleucus* (7,386, 11.6%), *Mus musculus* (6,245, 9.8%), *Arvicanthis niloticus* (5,497, 8.6%), *Mastomys huberti* (4,699, 7.4%) and Praomys daltoni (1,854, 2.9%) were the most commonly trapped.

Locations of points for the presence or absence of the seven most commonly trapped species are shown in Figure 3 (left column). These are able to identify locations of absence within an individual species documented range. Whether an absence represents a "true absence" (i.e. not present at this location) or a "pseudo absence" (i.e. not detected at this location) should be interpreted with consideration to the trapping effort at the site. This is compared to reported presence obtained from GBIF (centre column) and IUCN species range data (left column) in Figure 3.

```{r, figure_3, layout = "l-screen-inset", echo=FALSE}
knitr::include_graphics(here("figures", "Figure_3.png"))
```

@@

Sixty-two studies presented data on microorganisms that infect or are carried by small mammal species in West Africa. Seven studies solely investigated pathogens of rodents, including Hydatigera species (previously Taenia species) and Trichuris species. Fifty-five studies investigated microorganisms that were potentially zoonotic pathogens. Thirty-two microorganisms were tested for, 8 of these were defined at species level, with the remaining 24 at levels of higher taxonomic classification.

Thirty-two studies used Polymerase Chain Reaction (PCR) to detect the presence of 22 different species or families of microorganisms in 21,953 rodent samples. Eleven studies used antibody or antigen based molecular tests to detect the presence of 9 different species or families of microorganisms in 11,430 samples. Eight studies conducted histological or direct visualisation assays of samples for 11 parasitic or bacterial species in 11,229 samples. Three studies performed direct culture of Lassa mammarenavirus or Leishmania species to detect the presence of the pathogen in 643 samples. The most common microorganisms assessed for using PCR were Lassa mammarenavirus or other Arenaviridae (31%) the bacteria *Borrelia sp.* (11.3%) and *Bartonella sp.* (6.5%), followed by Usutu virus (6.4%) and Hantaviridae (5.6%). The most common pathogens assessed for with serology were *Lassa mammarenavirus* or other Arenaviridae (78%), *Toxoplasma gondii* (10.6%), *Borrelia sp.* (6.3%) and *Leptospirosis sp.* (2.2%). The most common pathogens assessed for with histology or direct visualisation were *Borrelia sp.* (48.7%), *Schistosoma sp*. (20.4%) and other parasites. All studies using direct culture were investigating either *Lassa mammarenavirus* (81%) or *Leishmania sp.* (19%).

Most studies tested for a single microorganisms (39), with 16 studies testing for two or more microorganisms. The most frequently tested for microorganisms were *Lassa mammarenavirus* (28%) or members of the Arenaviridae family (16%), the spirochete bacteria *Borrelia sp.* was investigated in 8 studies, *Bartonella sp.* and *Toxoplasma gondii* were investigated in 4 each, the remaining 25 microorganisms were reported in three or fewer studies.

There is spatial heterogeneity in the location of trapped rodents that were tested for the four most commonly assayed microorganisms. The presence and absence of Arenaviridae, predominantly *Lassa mammarenavirus*, in rodents was assessed in Guinea, Sierra Leone, Mali, Côte d'Ivoire, Benin and Nigeria. The known reservoir species of this zoonotic pathogen *Mastomys natalensis* was additionally detected in Senegal, Guinea-Bissau, Burkina-Faso and Niger with likely presence in Liberia and Togo. Arenaviridae were detected in rodents from 17 of 54 species or genera assayed from Guinea, Sierra Leone, Mali, Côte d'Ivoire, Ghana and Nigeria, no positive rodents were detected in Benin (Figure 4 top row). The presence and absence of *Borrelia sp.* was investigated in Senegal, Mauritania, Mali, Niger and Benin. Microorganisms of this genera were detected from samples in Mali with negative samples obtained from Senegal, Mauritania, Niger and Benin (Figure 4 second row). *Borrelia sp.* were detected in 8 of 44 species or genera of rodents assayed. The presence and absence of *Bartonella sp.* was investigated in Senegal, Mali and Nigeria with positive samples obtained at most sites from all three countries (Figure 4 third row). *Bartonella sp.* were detected in 11 of 11 specis or genera of rodents assayed. The presence of *Toxoplasma gondii* was only investigated in Senegal with relatively high proportions of positive samples at all sites (Figure 4 fourth row). *Toxoplasma gondii* was detected in 6 of 8 rodent species assayed.

```{r, figure_4, layout = "l-screen-inset", echo=FALSE}

@@

Ninety-seven rodent species were investigated for the presence of a zoonotic pathogen. 42,940 assays were performed on 32,014 individual rodents. The rodent species most commonly assayed for zoonotic pathogens were, *Rattus rattus* (n = 2,977) assessed for 24 pathogens, *Mus musculus* (n = 3,402) for 23, *Mastomys natalensis* (n = 7,189) and *Mastomys erythroleucus* (n = 3,013) both 19 and *Arvicanthis niloticus* (n = 3,840) for 18. All remaining species were investigated for 10 or fewer pathogens.

*Lassa mammarenavirus* was detected in 414 *Mastomys natalensis*, 9 *Mastomys erythroleucus*, 5 *Hylomyscus pamfi* and 3 *Mus baoulei* through PCR. In serological assays, evidence of prior infection or acquisition of maternal antibodies to *Lassa mammarenavirus* was detected in 465 *Mastomys natalensis*, 361 *Mastomys sp.*, 6 *Mus musculus* and 4 *Rattus rattus*. From direct viral culture 19 *Mastomys natalensis*, 2 *Rattus rattus* and 1 *Mus minutoides* were found to be infected with *Lassa mammarenavirus*. *Bartonella sp.* were detected in; 113 *Mastomys natalensis*, 50 *Mastomys erythroleucus*, 39 *Rattus norvegicus*, 16 *Rattus rattus* and 13 individuals of 7 further species through PCR.

*Borrelia sp.* infections were identified through PCR in 110 individuals; 37 *Mastomys erythroleucus*, 12 *Mus musculus*, 11 *Mastomys huberti* and *Praomys daltoni*. The remaining 39 positive results were from 9 further species. In serological assays, 47 *Mastomys natalensis*, 12 *Mastomys erythroleucus* and 8 *Praomys daltoni* were positive, with the remaining 11 positive individuals from three further species. Most *Borrelia sp.* infections were identified through histology or direct visualisation with, 243 *Arvicanthis niloticus*, 102 *Mastomys natalensis*, and 78 *Mastomys huberti* along with a further 64 individuals from 8 species infected.

*Toxoplasma gondii* was detected through PCR in 89 Mus musculus, 13 *Cricetomys gambianus*, 5 *Crocidura olivieri* and 3 *Rattus rattus*. From serological assays evidence of prior infection was found in 25 *Mus musculus*, 46 *Mastomys erythroleucus*, 4 *Rattus rattus* and *Mastomys natalensis*. A single *Mus musculus* and *Rattus rattus* were found to be infected using histology and direct visualisation. Finally, 38 species of Rodentia, 3 species of Soricomorpha and 2 species of Erinaceomorpha that were tested for potential pathogens had entirely negative results.

```{r figure_5, layout ="l-screen-inset", echo=FALSE}

@@

# Discussion

  * 124 trapping studies were identified across 14 West African countries. 54 studies provided additional information about potential zoonotic pathogens of trapped rodents.

>>>>

---
title: "A scoping review of rodent trapping studies to understand the prevalence of zoonotic disease in West Africa."
description: |
date: "`r Sys.Date()`"
author: 
  - first_name: "David"
    last_name: "Simons"
    url: https://www.dsimons.org
    orcid_id: 0000-0001-9655-1656
output: 
  distill::distill_article:
    toc: true
    toc_depth: 3
    toc_float: true
  # word_document:
  # pdf_document:
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_dir = here::here())
  })
always_allow_html: yes
bibliography: "`r here::here('citations', 'manuscript_citations.bib')`"
---
```{r, echo=FALSE}
source(here::here("scripts", "manuscript_library.R"))
```

@@

Rodents are important reservoirs of zoonotic infectious diseases; 255 species are known hosts of 85 zoonotic pathogens. These reservoir species are globally distributed with West Africa containing multiple known and several further predicted reservoir species. Typically International Union for Conservation of Nature (IUCN) species distribution maps and Global Biodiversity Information Facility (GBIF) presence maps are used to determine regions at risk for zoonotic spillover events, however, these datasets are biased by incomplete sampling. To investigate the impact of rodent species sampling heterogeneity on zoonotic spillover risk we have systematically reviewed rodent trapping studies to produce a contextually rich dataset that can be used to explore this risk at finer spatial resolution. Here we show that sampling of rodents and their pathogens in the West African region are spatially biased by country and trapping habitat. We found that rodent trapping effort was associated with regional population density and was increased in habitats modified by human activity. We produce updated rodent species ranges compared to available IUCN maps and enrich GBIF data by including locations of rodent species absence and trapping effort. Furthermore, we report the spatial bias in the investigation of four important rodent zoonoses, Arenaviridae, *Borellia sp.*, *Bartonella sp.* and *Toxoplasma sp.*. Our results highlight that incorporating sampling bias is important when assessing the risk for zoonotic spillover events from rodents. The synthesis of contextually rich rodent trapping data contributes important information that is lacking in IUCN distribution maps and GBIF species presence data. We anticipate this dataset can support the production of more complete spatial risk assessments of zoonotic spillover events. For example, the inclusion of absence data and trapping effort can help identify regions where data sparsity may produce inappropriately predicted risk. Furthermore, these data identify regions in need of comprehensive study for our four targeted rodent zoonoses, Arenaviridae, *Borellia sp.*, *Bartonella sp.* and *Toxoplasma sp.*.

@@

The potential effect of zoonotic infectious diseases outbreaks on human health and societies has been dramatically highlighted through the ongoing SARS-CoV-2 pandemic and recent Ebola virus outbreaks. The number of zoonotic spillover events are projected to increase under intensifying anthropogenic pressure such as, increased human populations [@allen_global_2017], increasing urbanisation [@hassell_urbanization_2017], and global climate change [@morse_prediction_2012]. In addition, two taxa are proposed to contribute to the greatest number of potential zoonotuc pathogens globally, rodents (Rodentia) and bats (Chiroptera) [@han_rodent_2015]. Of 2,220 extant rodent species, 244 (10.7%) are described as being reservoirs of 85 zoonoses [@han_global_2016]. Specifically, West Africa has previously been identified as region at increased hazard for rodent-borne zoonotic spillover events under different projected scenarios [@garcia-pena_land-use_2021]. Rodents are implicated in the transmission pathways of several human diseases in this region, including, Lassa fever, Schistosomiasis and Leptospirosis.

Rodent species form diverse assemblages, which provide important and beneficial ecosystem services including pest regulation and seed dispersal [@fischer_ecosystem_2018]. The role of rodent species' in zoonotic infectious disease spillover or novel pathogen emergence can be viewed as an ecosystem disservice. Rodents typically demonstrate “fast” life histories [@dobson_fast_2007], these traits are also associated with a species being a reservoir of zoonotic pathogens [@han_rodent_2015; @albery_fast-lived_2021]. Further, these traits are prevalent in species that thrive in human dominated landscapes where they displace species that are less likely to be reservoirs of zoonotic pathogen [@gibb_zoonotic_2020]. The widespread occurrence of reservoir species and their proximity to human activity make the description of rodent species assemblages and host-pathogen interactions vitally important to understanding the risk of zoonotic spillover and novel pathogen emergence.

Rodent trapping studies provide a useful method to describe rodent population structures and survey for potential zoonoses. Studies have been conducted in West Africa to both identify novel potential zoonotic pathogens within rodents [@usaid_one_2021] and to investigate the prevalence and burden of known pathogens within known rodent hosts (e.g. for Lassa fever [@fichet-calvet_diversity_2009] and Schistosomiasis [@catalano_multihost_2020]). These studies provide contextually rich information on when, where and under what conditions rodents were trapped that are typically missing from the global datasets [@bovendorp_optimising_2017]. Despite this, these global datasets (e.g. IUCN, GBIF and GIDEON) are typically used to inform species distribution maps and zoonotic disease spillover predictions [@han_global_2016; @smith_global_2014; @pigott_mapping_2014].

These studies have been used to identify potential geographic hotspots where virus and host species diversity may be expected to be at its greatest to predict regional zoonotic disease spillover risks. However, there remains the potential for important confounding in these spatial distributions through bias generated by study design and selection of sampling sites. For example, systematically increased sampling (e.g. more intensive studies over longer time periods) or over-representation of certain habitats (e.g. peri-urban landscapes) could lead to an apparent association between locations and risk that is driven by these factors rather than an underlying host and virus association [@wille_how_2021; @gibb_mammal_2021]. Conversely some regions may not be sampled adequately and therefore under-represented in these datasets due to sparse human populations or inaccessible habitats leading to these regions being reported to be at low risk of zoonotic disease spillover events.

@@

For studies investigating rodent zoonoses we compared the location of trapping sites with SEDAC Global Population Density estimates for 2005 - the median year studies were commenced. We used a Generalised Additive Model (GAM) incorporating a spatial interaction term to investigate the association of number of trap nights and human population density. The model structure was specified as:

@@

We identified 4,282 relevant citations, with 126 rodent trapping studies included. The earliest included studies were published in 1974 with increasing numbers of studies being performed annually since 2005. The median length of rodent trapping activity was one year (IQR 0-2 years). The median time from completion of rodent trapping to publication is 3 years (IQR 2-5 years) (Figure 1.).

@@

if(knitr::is_html_output()) {
  knitr::include_graphics(here("figures", "Fig_1_Panel_A.png"), dpi = 320)
} else {
  knitr::include_graphics(here("figures", "Fig_1_Panel_A.png"))
}

@@

Rodent trapping took place at 1,331 trap sites with at least one trap site recorded from 14 West African countries. No rodent trapping studies were identified from Gambia or Togo. The most frequently studied countries were Senegal, Ghana, Guinea and Nigeria (more than 10 studies each). Fewer than 5 studies were conducted in Liberia, Burkina Faso and Guinea-Bissau. Thirty-one (25%) studies reported trapping at a single study site, 46 (37%) studies trapped at between two and five study sites, the remaining 49 studies trapped at between six and 93 study sites. Trap sites were situated in 273 (19.3%) of 1414 level 2 administrative regions in the 14 West African nations. The areas with highest trap night density included the capital cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako) and Ghana (Accra) and the largest cities of Côte d'Ivoire (Abidjan) and Benin (Cotonou). Outside of these cities, Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone had the highest density of trap nights (Figure 2. A).

Greater numbers of trap nights were conducted in areas with higher population densities (*R~2~* = 0.18) adjusting for spatial autocorrelation. In particular Mauritania, Northern Senegal, Eastern Sierra Leone, Eastern Guinea and South West Nigeria were trapped at higher rates than would be expected based on their population densities. South West Nigeria, Northern Nigeria, Liberia, Côte d'Ivoire, Ghana, Niger and Burkina Faso were trapped at lower rates than would be expected.

@@

```{r, figure_2, layout = "l-screen-inset", echo=FALSE}

knitr::include_graphics(here("figures", "Figure_2.png"))

@@

73,164 small mammals were trapped (592 were trapped outside of West African countries), 2,830 (4%) trapped individuals were identified to order level (Rodentia), 7,760 (11%) were identified to genus level, the remaining 62,574 (85%) were identified to species level. In studies reporting the number of trap nights the median trap success rate of traps placed in or around buildings was 13% (IQR 6-24%), this compares to a median trap success rate of 3% (IQR 1-9%) in other habitats. The majority of the 147 species trapped were Rodentia (112) (Supplementary Material 3), of these Muridae (82) were the largest family of rodents, followed by, Sciuridae (10), Gliridae and Nesomyidae (both 6), Ctenodactylidae, Anomaluridae, Dipodidae, Hystricidae and Thryonomidae (2 or fewer). The remaining 34 species were from the orders of Soricomorpha (30), Erinaceomorpha (2) and Afrosoricida (1).

*Mastomys sp.* were the most commonly trapped genus (27,079, 37.7%), followed by, *Rattus sp.* (11,472, 16.1%), *Mus sp.* (8,624, 12.0%), *Arvicanthis sp.* (5,821, 8.1%) and *Praomys sp.* (5,409, 7.5%). *Mastomys natalensis* were the most commonly trapped species (11,116, 17.4%), followed by, *Rattus rattus* (9,959, 15.6%), *Mastomys erythroleucus* (7,386, 11.6%), *Mus musculus* (6,245, 9.8%), *Arvicanthis niloticus* (5,497, 8.6%), *Mastomys huberti* (4,699, 7.4%) and *Praomys daltoni* (1,854, 2.9%).

Locations of points for the presence or absence of the seven most commonly trapped species are shown in Figure 3 (left column). Absence include true absences and pseudo-absences dependent on the trapping effort at the site. This are shown compared to species presence obtained from GBIF (centre column) and IUCN species range data (left column) (Figure 3).

```{r, figure_3, layout = "l-screen-inset", echo=FALSE}

knitr::include_graphics(here("figures", "Figure_3.png"))

@@

Sixty-two studies presented data on microorganisms that infect or are carried by small mammal species in West Africa. Seven studies solely investigated pathogens of rodents, including Hydatigera species (previously Taenia species) and Trichuris species. Zoonotic pathogens and potentially zoonotic micro-organisms were investigated in 55 studies. Thirty-two microorganisms were tested for, 8 of these were specified to species level, the remaining 24 were specified at higher taxonomic classifications.

Polymerase Chain Reaction (PCR) was used in 32 studies to detect the presence of 22 different species or families of microorganisms from 21,953 rodent samples. The most common microorganisms assayed for using PCR were *Lassa mammarenavirus* or other Arenaviridae (31%) the bacteria *Borrelia sp.* (11.3%) and *Bartonella sp.* (6.5%), followed by Usutu virus (6.4%) and Hantaviridae (5.6%). Molecular tests (antibody or antigen based) were used in 11 studies to detect the presence of 9 different species or families of microorganisms from 11,430 rodent samples. The most common pathogens assessed for with serology were *Lassa mammarenavirus* or other Arenaviridae (78%), *Toxoplasma gondii* (10.6%), *Borrelia sp.* (6.3%) and *Leptospirosis sp.* (2.2%). Histological or direct visualisation assays were used in 8 studies investigating 11 parasitic or bacterial species from 11,229 rodent samples. The most common pathogens assessed for with histology or direct visualisation were *Borrelia sp.* (48.7%), *Schistosoma sp*. (20.4%) and other parasites. Direct culture of *Lassa mammarenavirus* or Leishmania species were reported from 3 studies to detect the presence of these pathogen in 643 rodent samples.

Most studies tested for a single microorganisms (39), 16 studies tested for two or more microorganisms. The most frequently investigated microorganisms were *Lassa mammarenavirus* (17 studies) or members of the Arenaviridae family (10), the spirochete bacteria *Borrelia sp.* (8), *Bartonella sp.* and *Toxoplasma gondii* (4), the remaining 25 microorganisms were reported in three or fewer studies.

The location of trapped rodents that were tested for the four most commonly assayed microorganisms did not encompass the entire range of these hosts. The presence and absence of Arenaviridae, predominantly *Lassa mammarenavirus*, in rodents was investigated in Benin, Côte d'Ivoire, Guinea, Mali, Nigeria and Sierra Leone. No studies investigated the prevalence of these pathogens in hosts from Burkina-Faso, Guinea-Bissau, Liberia, Niger, Senegal and Togo. Arenaviridae were detected in rodents from 17 of 54 species or genera assayed in all the countries it was investigated except Benin (Figure 4 top row). The presence and absence of *Bartonella sp.* was investigated in Mali and Nigeria and Senegal with positive samples detected in 11 of 11 species or genera of rodents obtained at most sites from these countries (Figure 4 second row). The presence and absence of *Borrelia sp.* was investigated in Benin, Mali, Mauritania, Niger and Senegal. *Borrelia sp.* were detected in 8 of 44 species or genera of rodents assayed in Mali with negative samples obtained from Benin, Mauritania, Niger and Senegal (Figure 4 third row). The presence of *Toxoplasma gondii* was only investigated in Senegal and was detected in 6 of 8 rodent species assayed (Figure 4 fourth row).

```{r, figure_4, echo=FALSE}

@@

Ninety-seven rodent species were investigated for the presence of a zoonotic pathogen, 42,940 assays were performed on 32,014 individual rodent samples. The rodent species most commonly assayed for zoonotic pathogens were, *Rattus rattus* (n = 2,977) assessed for 24 pathogens, *Mus musculus* (n = 3,402) for 23, *Mastomys natalensis* (n = 7,189) and *Mastomys erythroleucus* (n = 3,013) both 19 and *Arvicanthis niloticus* (n = 3,840) for 18. All remaining species were investigated for 10 or fewer pathogens.

*Lassa mammarenavirus* was detected in 14 species of rodents and Crocidura genera. *Mastomys natalensis* was the most common species to test positive (Figure 4.). Fifty-four species and genera were tested for arenaviruses, with 37 testing negative. *Bartonella sp.* infections were identified in all 11 species of rodents and shrews that were tested. *Mastomys erythroleucus* was the most common species to test positive. *Borrelia sp.* infections were identified in 13 species of rodents and shrews. *Mastomys natalensis* was the most common species to test positive. Forty-four species and genera were tested for *Borrelia sp.* with 36 testing negative. *Toxoplasma gondii* was detected in 6 species of rodents and shrews, with all individuals from two species *Arvicanthis niloticus* and *Praomys daltoni* testing negative. *Mus musculus* was the most common species to test positive. Finally, 38 species of Rodentia, 3 species of Soricomorpha and 2 species of Erinaceomorpha that were tested for pathogens had entirely negative results.

```{r figure_5, echo=FALSE}

@@

# Discussion    
We identified 124 rodent trapping studies conducted in the West African region since 1965. These studies were conducted in 14 countries with 54 studies providing information about zoonotic pathogens in trapped rodents. Studies were conducted in all West African countries except Gambia and Togo with between- and within-country variability in the number of trap nights conducted. Over 70,000 rodents identified to genus or species level were trapped from 112 species of Rodentia. Thirty-two potential pathogens were investigated across all studies, 5 rodent species were assayed for at least 18 potential pathogens. We identify several important limitations to available data, potentially biasing previous zoonotic risk studies, first, in the timing of studies, and the locations in which they've been conducted, second, the locations of species presence and third, the selection of pathogens to assay.

First, timing and location of trapping studies can introduce bias to inference from these data when investigating the risk of zoonotic disease emergence. Most studies have been conducted within the last twenty years. Human population, intensification of agriculture and urbanisation have proceeded rapidly over this time across West Africa. These changes modify the available habitats for rodent species increasing the prevalence of commensal species that are more likely to be reservoirs of zoonoses. Not accounting for the time period in which data were collected will introduce bias into inferences made for zoonotic emergence based on reservoir species distributions. Further, several regions, namely North West Senegal, Eastern Sierra Leone, Southern Ghana, Southern Benin, South West Nigeria  Central Nigeria have been sampled at much higher intensity than elsewhere in West Africa. Studies modelling risk of zoonoses without incorporating this important sampling heterogeneity will introduce bias when inferring current or projected risk.

Second, rodent species were found to have wider distributions than the ranges reported by IUCN data. Trapping studies were conducted from more regions across West Africa than were available from GBIF records and where therefore able to report presence and absence of species at greater spatial resolution. Further, the additional data produced from species' absence and pseudoabsence highlighted the variability of rodent presence within their ranges across the region. Failing to incorporate this heterogeneity in rodent reservoir species' distribution can importantly bias estimations of the risk of zoonotic disease emergence. For example, the apparent absence of *Mastomys natalensis* at several trapping sites in Nigeria may explain some of the observed spatial variation in Lassa fever spillover that would not be expected based on it's IUCN proposed range or presence only data from GBIF.

Third, while up to 32 potential zoonoses were investigated only four were investigated in more than three studies. Of these four studies conducting assays for *Lassa mammarenavirus* or other Arenaviridae where the most widely conducted in regards to number of countries and number of study sites within countries. For example, *Toxoplasma sp.* were found to be prevalent in rodent hosts within Senegal and despite a significant burden of human disease no studies investigated the prevalence of this pathogen in other countries of West Africa [https://www.nature.com/articles/s41598-020-69078-9]. The limited geographic scope of sampling for these pathogens hampers efforts to understand the risk of zoonotic spillover across this heterogeneous region.

Finally, we found that only five of the 112 rodent species trapped had been tested for more than half of the potential pathogens. There is increasing emphasis on the role of co-infection and multi-species pathogen transmission systems on zoonotic disease spillover risk which depends on comprehensive testing of both known and un-known reservoir species. More comprehensive testing of potential reservoir species across West Africa is required to better understand the role of individual species in disease systems. This will also reduce the risk of bias in predictive risk modelling by better characterising host-pathogen associations in species rich environments.

## Implications

## Limitations

## Future research

# Conclusion

# Notes

First, we found that study locations were biased towards regions with higher human population densities and were more likely to be conducted within habitats associated with anthropogenic land use change. Second, we were able to obtain contextually rich information on the presence and absence of 147 small-mammal species to compare to IUCN and GBIF distributions and investigate spatial bias in species distributions in these datasets. Third, we synthesised data on the presence and absence of zoonotic pathogens detected within small-mammals from these studies and identified biases in the spatial and species sampling across West Africa. Finally, we synthesise host-pathogen associations from the included studies.

## Spatial and habitat biases in rodent sampling

Between countries there were large differences in the number of studies performed in each country and the intensity of trapping. Important heterogeneity was observed within countries with most countries sparsely sampled, Senegal and Sierra Leone were most comprehensively sampled with studies conducted in all but 7 level-2 administrative regions.

We found a non-linear positive relationship between human population density and the trap-night density of studied regions, suggesting that studies were preferentially conducted in regions of higher human habitation. Further, we found that regions consisting of habitat types associated with anthropogenic disturbance (e.g. mosaic vegetation, cropland) were more commonly sampled. Together, these results suggest that rodent sampling in West Africa is limited by biases introduced by the selection of study sites that impacts the inference and generalisability of results. This can importantly limit the applicability of out-of-sample predictions generated from models dependent on these data.

## Rodent presence and absence

Included studies trapped over 70,000 rodents identified to genus or species level, 112 species of Rodentia were reported. The most commonly trapped species were the commensal rodents *Mastomys natalensis*, *Rattus rattus*, *Mastomys erythroleucus* and *Mus musculus*. Rodents were more frequently trapped in or around buildings (13% trap success) compared with other habitats (3%). This difference potentially explains some of the observed bias in habitat selection for studies as more rodents are trapped for the same trapping effort. However, this may also suggest that rodents residing in human dominated landscapes are less "trap-shy" or that there is an increased abundance of rodents in these settings. Together these factors complicate comparisons made between urban and non-urban studies.
